On probabilistic one-vs-one classification

### 2025-04-23 paper.tex
Added proof for proposition 3

\emph{Canonicity} refers to requirement that the method should be in a natural way unique. Borrowing an example from regression analysis, Gauss-Markov theorem states that ordinary least squares (OLS) estimates is best linear unbiased estimate of the coefficients of a linear model. Therefore OLS  is a \emph{canonical} way to estimate parameters of a regression model. Note that potentially there may be multiple canonical methods, derived from (or satisfying) differing sets of assumptions.

@@

Let us make this explicit. Let $\boldsymbol{v}$ be a coupling method. Denote by  $\pi$ the change of priors $\pi:\Pi \rightarrow \Pi'$. If we have a  binary classifier predicting distribution $D \propto (p, p')$ on two classes $C_i$ and $C_j$, then we define new distribution $\pi_2^{ij}(D)$ by requiring
$$
\pi_2^{ij}(D) \propto \biggl(p \frac{P'_i}{P_i},p' \frac{P'_j}{P_j}\biggr).
$$
We denote by $\boldsymbol{R}^\pi$ the matrix with entries $\pi_2^{ij}(r_{ij})$. The value $\boldsymbol{v}(\boldsymbol{R}^\pi)$ represents effect of change of priors if we apply \eqref{eq:changePrior} \emph{before} coupling.

On the other hand we can apply \eqref{eq:changePrior} after coupling. Say we have a distribution on $K$ classes
\begin{align}
D \propto (p_1, \ldots, p_K)
\end{align}
then we define $D^\pi$ by requiring 
\begin{align}
D^\pi \propto \biggl(p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}\biggr).
\end{align}
We now say that a method if Bayes covariant if
\begin{align}
\boldsymbol{v}(\boldsymbol{R}^\pi)=	\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi,
\end{align}
for any pairwise predictions satisfying Assumption \ref{ass:1}.

@@

\begin{prop}
Any arboreal method is Bayes covariant.
\end{prop}
\begin{proof}
	We can proceed by induction. The statement is clear when $K=2$. Suppose the proposition is true for an integers $<K$ and we are given a spanning tree $T$ on $K$ vertices. Any tree has a vertex of degree one. We may assume that in the case of $T$ it is the vertex corresponding to $C_1$ and the unique edge incident to $C_1$ is $C_1C_2$.

	 Denote by $\boldsymbol{t}$ the prediction corresponding to $T$ and by $\boldsymbol{t}'$ the prediction corresponding to the subgraph without the vertex $C_1$. Let us write 
	\begin{align*}
	\boldsymbol{t}(x) &= (t_1, t_2, \ldots, t_K) \\
	\boldsymbol{t'}(x) &= ( t'_2, \ldots, t'_K) 
\end{align*}
These vectors are related as follows:
\begin{align}
\begin{split}
\frac{t_1}{t_1+t_2} &= r_{12} \\
(t_2, \ldots, t_K) &\propto (t'_2, \ldots, t'_{K})
\end{split} \label{eq:ind1}
\end{align}

Now consider the effect of change of priors $\Pi = (P_1, \ldots, P_K)$ to $\Pi' = (P'_1, \ldots, P'_K)$. Applying \eqref{eq:changePrior} to the multi-class prediction yields a vector 
$$
\boldsymbol{t}^{\Pi\rightarrow \Pi'} \propto (t_1 \frac{P'_1}{P_1}, \ldots, t_K \frac{P'_K}{P_K}).
$$
Now consider what happens when we apply \eqref{eq:changePrior} before coupling. By induction assumption the method $\boldsymbol{t}'$ is Bayes covariant and thus applying coupling \emph{after} change of priors yields
$$
\boldsymbol{t}'^{\Pi\rightarrow \Pi'} \propto (t_2 \frac{P'_2}{P_2}, \ldots, t_K \frac{P'_K}{P_K})
$$

@@

Suppose real numbers $\alpha_1, \ldots,\alpha_L$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_L$ are probabibilistic multi-class classification methods satisfying Assumption \ref{ass:1}. We define their linear combination $\bigoplus_i a_i \boldsymbol{v}_i$ as the classification method that for any $x$ yields the probability distribution on $L$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_L \boldsymbol{v}_L\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_L^{a_L}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication. With this definition we can state the following result.

@@

\begin{proof}

>>>>

\emph{Canonicity} refers to requirement that the method should be natural. Borrowing an example from regression analysis, Gauss-Markov theorem states that ordinary least squares (OLS) estimates is best linear unbiased estimate of the coefficients of a linear model. Therefore OLS  is a \emph{canonical} way to estimate parameters of a regression model.

Note that potentially there may be multiple canonical methods, derived from (or satisfying) differing sets of assumptions. For instance, in the theory of scoring rules, both logarithmic and quadratic scores can be viewed as canonical [\cite{shannon1948mathematical,selten1998axiomatic}].

@@

Let us make this explicit. Let $\boldsymbol{v}$ be a coupling method. Denote by  $\pi$ the change of priors $\pi:\Pi \rightarrow \Pi'$. If we have a  binary classifier predicting distribution $D \propto (p, p')$ on two classes $C_i$ and $C_j$, then we define new distribution $\pi_2^{ij}(D)$ by requiring
$$
\pi_2^{ij}(D) \propto \biggl(p \frac{P'_i}{P_i},p' \frac{P'_j}{P_j}\biggr).
$$
We denote by $\boldsymbol{R}^\pi$ the $K\times K$ matrix with entries $\pi_2^{ij}(r_{ij})$. The value $\boldsymbol{v}(\boldsymbol{R}^\pi)$ represents the effect of change of priors if we apply \eqref{eq:changePrior} \emph{before} coupling.

On the other hand we can apply \eqref{eq:changePrior} after coupling. Say we have a distribution on $K$ classes
\begin{align}
D \propto (p_1, \ldots, p_K)
\end{align}
then we define $D^\pi$ by requiring 
\begin{align}
D^\pi \propto \biggl(p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}\biggr).
\end{align}
We now say that a method if Bayes covariant if
\begin{align}
\boldsymbol{v}(\boldsymbol{R}^\pi)=	\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi,
\end{align}
for any matrix $\boldsymbol{R}$ with $r_{ij}> 0$.

@@

\begin{prop}
Any arboreal method $\boldsymbol{t}$ is Bayes covariant.
\end{prop}
\begin{proof}
We start by defining property $BC_{ij}$ (short for Bayes covariant when restricted to classes $i$ and $j$) of a coupling method. A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal.

Let us list some easy to verify properties of this concept:

\begin{itemize}
\item[a)] a Bayes covariant coupling method satisfies $BC_{ij}$ for any $i\not= j$,
\item[b)] if a coupling method satisfies $BC_{12}, BC_{23}, \ldots, BC_{(K-1),K}$ then it is Bayes covariant,
\item[c)] if a coupling method satisfies $BC_{ij}$ for all $i\not= j$ then it is Bayes covariant,
\item[d)] if a coupling method satisfies $BC_{ij}$ and $BC_{jk}$ then it satisfies $BC_{ik}$.
\end{itemize}

An arboreal coupling method $\boldsymbol{t}$ induced by a spanning tree $T$ on the classes satisfies $BC_{ij}$ whenever the edge $C_iC_j$ belongs to $T$. It follows using from b) and d) that the method is Bayes covariant.

%	
%	
%Let $\pi$ be a change of priors from $\Pi = (P_1, \ldots, P_K)$ to $(P'_1, \ldots, P'_K)$. We need to verify $\boldsymbol{v}(\boldsymbol{R}^\pi) =\bigl( \boldsymbol{v}(\boldsymbol{R})\bigr)$ for any pairwise matrix $\boldsymbol{R}$. |We will rephrase this in an equivalent way. Let 
%\begin{align}
%	\begin{split}
%	\boldsymbol{v}(\boldsymbol{R}^\pi) &= (v'_1, \ldots, v'_K) \\
%	\boldsymbol{v}(\boldsymbol{R}) &= (v_1, \ldots, v_K)
%	\end{split}
%\end{align}
%Suppose $\pi$ is a change of priors from $\Pi = (P_1, \ldots, P_K)$ to $\Pi' = (P'_1, \ldots, P'_K)$. If the prediction given $\Pi$ is $(t_1, \ldots, t_K)$ then we need to prove that
%\begin{align*}
%\boldsymbol{t}(R^\pi) = \bigl(t_1 \frac{P'_1}{P_1}, \ldots, t_K \frac{P'_K}{P_K}\bigr)
%\end{align*}
%	
%	We can proceed by induction. The statement is clear when $K=2$. Suppose the proposition is true for an integers $<K$ and we are given a spanning tree $T$ on $K$ vertices. Any tree has a vertex of degree one. We may assume that in the case of $T$ it is the vertex corresponding to $C_1$ and the unique edge incident to $C_1$ is $C_1C_2$.
%	
%	 Denote by $\boldsymbol{t}$ the prediction corresponding to $T$ and by $\boldsymbol{t}'$ the prediction corresponding to the subgraph without the vertex $C_1$. Let us write 
%	\begin{align*}
%	\boldsymbol{t}(x) &= (t_1, t_2, \ldots, t_K) \\
%	\boldsymbol{t'}(x) &= ( t'_2, \ldots, t'_K) 
%\end{align*}
%These vectors are related as follows:
%\begin{align}
%\begin{split}
%\frac{t_1}{t_1+t_2} &= r_{12} \\
%(t_2, \ldots, t_K) &\propto (t'_2, \ldots, t'_{K})
%\end{split} \label{eq:ind1}
%\end{align}
%
%Now consider the effect of change of priors $\Pi = (P_1, \ldots, P_K)$ to $\Pi' = (P'_1, \ldots, P'_K)$. Applying \eqref{eq:changePrior} to the multi-class prediction yields a vector 
%$$
%\boldsymbol{t}^{\Pi\rightarrow \Pi'} \propto (t_1 \frac{P'_1}{P_1}, \ldots, t_K \frac{P'_K}{P_K}).
%$$
%Now consider what happens when we apply \eqref{eq:changePrior} before coupling. By induction assumption the method $\boldsymbol{t}'$ is Bayes covariant and thus applying coupling \emph{after} change of priors yields
%$$
%\boldsymbol{t}'^{\Pi\rightarrow \Pi'} \propto (t_2 \frac{P'_2}{P_2}, \ldots, t_K \frac{P'_K}{P_K})
%$$

@@

Suppose real numbers $\alpha_1, \ldots,\alpha_L$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_L$ are probabibilistic multi-class classification methods satisfying Assumption \ref{ass:1}. We define their linear combination $\bigoplus_i a_i \boldsymbol{v}_i$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_L \boldsymbol{v}_L\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_L^{a_L}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

@@

\begin{proof}
We will be using property $BC_{ij}$ from the proof of the previous proposition. It is sufficient to show that $BC_{ij}$ holds for any $i\not= j$.

Let $\pi$ be a change of priors from $(P_1, \ldots, P_K)$  to $(P'_1, \ldots, P'_K)$. Let $o_l$ denotes the odds of classes $C_i$ and $C_j$ predicted by coupling method $\boldsymbol{v}_l$. Then the odds of $C_i$ and $C_j$ predicted by $(\bigoplus_i a_i \boldsymbol{v}_i)^\pi$ are 
\begin{align}
	\biggl(\bigoplus_i a_i \boldsymbol{v}_i\biggr)^\pi(\boldsymbol{R}) & = \frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggl(\bigoplus_i a_i \boldsymbol{v}_i\biggr)(\boldsymbol{R}) \\
	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_l o_l^{a_l}
\end{align}

Let us now use Bayes covariant of $\boldsymbol{v}_i$. We have
\begin{align}
	\biggl(\bigoplus_i a_i \boldsymbol{v}_i\biggr)(\boldsymbol{R^\pi}) & =  \biggl(\bigoplus_i a_i \boldsymbol{v}_i^\pi\biggr)(\boldsymbol{R}) \\
	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_l o_l^{a_l}
\end{align}

It follows that the odds of $C_i$ to $C_j$ for $\bigoplus_i a_i \boldsymbol{v}_i$ are 
\begin{align}
\biggl(\frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggr)^{\sum a_i} \prod_l o_l^{a_l}.
\end{align}
Clearly, under the assumption $\sum_i a_i = 1$, the classifier $\bigoplus_i a_i \boldsymbol{v}_i$ satisfies $BC_{ij}$ and it follows that it is Bayes covariant.
\end{proof}
